### 2017-07-18 01-intro.tex
Process comments by Kees V and Ceriel (but don't oblige some of them)

\section{Introduction}
The tremendous amount of data that we generate through our daily applications such as
social networking services, online shopping, and news recommendations, provides us
with an opportunity to extract hidden but useful, even invaluable information.
Realizing this opportunity, however, requires a significant amount of effort because
traditional machine learning algorithms may become inefficient with large amounts of data.

@@

One such big data problem is analyzing large graphs such as social networks where it is not unusual to see a network consisting of billions of edges and tens of millions of vertices~\cite{yang2015defining}. In this paper, we address the overlapping community detection problem~\cite{xie2013overlapping}, where the goal is to learn the probability distribution of each vertex to participate in each community, given a set of vertices, the links between them (which are usually very sparse), and the number of latent communities. A community can be seen as a densely connected group of vertices that are only sparsely connected to the rest of the network. This problem is significantly more complex than the related domain of detecting disjoint communities.

@@

Starting from an existing implementation in Python/Numpy, we create an efficient
single-threaded C++ version to serve as the baseline for the parallel versions.
This already improves performance by a factor of 1000 to 1500 compared to
Python.
% This transformation is not considered one of our key contributions.
Then, our key contributions to the algorithm are:

>>>>

\section{Introduction}
The tremendous amount of data that we generate through our daily applications such as
social networking services, online shopping, and news recommendations, provides us
with an opportunity to extract hidden but useful, even invaluable information.
Realizing this opportunity, however, requires significant efforts because
traditional machine learning algorithms may become inefficient with large amounts of data.

@@

One big data problem is analysis of large graphs such as social networks where it is not unusual to see a network consisting of billions of edges and tens of millions of vertices~\cite{yang2015defining}. In this paper, we address the overlapping community detection problem~\cite{xie2013overlapping}, where the goal is to learn the probability distribution of each vertex to participate in each community, given a set of vertices, the links between them (which are usually very sparse), and the number of latent communities. A community can be seen as a densely connected group of vertices that are only sparsely connected to the rest of the network. This problem is significantly more complex than the related domain of detecting disjoint communities.

@@

Starting from an existing implementation in Python/Numpy, we created an efficient
single-threaded C++ version to serve as the baseline for the parallel versions.
This already improves performance by a factor of 1000 to 1500 compared to
Python.
% This transformation is not considered one of our key contributions.
Then, our key contributions to the algorithm are:

### 2017-07-18 02-01-background.tex
Process comments by Kees V and Ceriel (but don't oblige some of them)

\looseness=-1
Stochastic gradient Riemannian Langevin dynamics (SGRLD)~\cite{patterson2013stochastic} is a subclass of SGLD which is developed to efficiently sample from the probability simplex. By applying Riemannian geometry~\cite{girolami2011riemann} and using the mini-batch-based estimator in Eq.~\ref{eqn:sgld_update}, it achieved state-of-the-art performance for latent Dirichlet allocation (LDA)~\cite{blei2003latent}. In particular, for a $K$-dimensional probability simplex $\pi$, it uses the \textit{expanded-mean} re-parameterization trick, where the probability of a category $k$ is given by $\pi_k = \ta_k / \sum_{j=1}^K \ta_j$ with $\ta_k \sim \text{Gamma}(\al,1)$ and $\al$ a hyperparameter of the Dirichlet distribution $p(\pi | \al )$. Then, the update rule of SGRLD is:
\bea
\ta_{k}^* \law \left| \ta_{k} + \f{\ep_t}{2} \left( \al - \ta_{k} + \f{N}{|\SGLDMinibatch|} \sum_{x \in \SGLDMinibatch} g_x(\ta_{k}) \right) + (\ta_{k})^\ha \xi_{t} \right|, \label{eqn:sgrld_update}
\eea
here $g_x(\ta_{k})$ is the gradient of the log posterior w.r.t. $\ta_{k}$ on a data point $x\in \SGLDMinibatch$.

@@

\paragraph*{\textbf{Sampling local parameters}}
the update rule for the local parameters $\phi$ is
\bea
\phi_{ak}^* \law \left| \phi_{ak} + \f{\ep}{2} \left( \al - \phi_{ak} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} g_{ab}(\phi_{ak})\right) \right. \nn \\ \left. + (\phi_{ak})^{\ha} \xi_{ak}\right|,
\label{eqn:local_update}
\eea
where the gradient w.r.t. $\phi$ is
\bea
g_{ab}(\phi_{ak}) = \f{f_{ab}^{(y)}(k)}{Z_{ab}^{(y)}\phi_{ak}} - \f{1}{\sum_j\phi_{aj}}.
\eea
Here, $\Neighbors$ is the neighbor set for a node in the mini-batch $\Minibatch$. $\Neighbors$ is another mini-batch of nodes sampled from \Vertices. Note that $|\Neighbors| \ll |\Vertices|=N$.

>>>>

Stochastic gradient Riemannian Langevin dynamics (SGRLD)~\cite{patterson2013stochastic} is a subclass of SGLD which is developed to efficiently sample from the probability simplex. By applying Riemannian geometry~\cite{girolami2011riemann} and using the mini-batch-based estimator in Eq.~\ref{eqn:sgld_update}, it achieved state-of-the-art performance for latent Dirichlet allocation (LDA)~\cite{blei2003latent}. In particular, for a $K$-dimensional probability simplex $\pi$, it uses the \textit{expanded-mean} re-parameterization trick, where the probability of a category $k$ is given by $\pi_k = \ta_k / \sum_{j=1}^K \ta_j$ with $\ta_k \sim \text{Gamma}(\al,1)$ and $\al$ a hyperparameter of the Dirichlet distribution $p(\pi | \al )$. Then, the update rule of SGRLD is:
\bea
\ta_{k}^* \law \left| \ta_{k} + \f{\ep_t}{2} \left( \al - \ta_{k} + \f{N}{|\SGLDMinibatch|} \sum_{x \in \SGLDMinibatch} g_x(\ta_{k}) \right) + (\ta_{k})^\ha \xi_{t} \right|, \label{eqn:sgrld_update}
\eea
here $g_x(\ta_{k})$ is the gradient of the log posterior w.r.t. $\ta_{k}$ on a data point $x\in \SGLDMinibatch$.

@@

\paragraph*{\textbf{Sampling local parameters}}
the update rule for the local parameters $\phi$ is
\bea
\phi_{ak}^* \law \left| \phi_{ak} + \f{\ep}{2} \left( \al - \phi_{ak} + \f{N}{|\Neighbors|} \sum_{b \in \Neighbors} g_{ab}(\phi_{ak})\right) \right. \nn \\ \left. + (\phi_{ak})^{\ha} \xi_{ak}\right|,
\label{eqn:local_update}
\eea
where the gradient w.r.t. $\phi$ is
\bea
g_{ab}(\phi_{ak}) = \f{f_{ab}^{(y)}(k)}{Z_{ab}^{(y)}\phi_{ak}} - \f{1}{\sum_j\phi_{aj}}.
\label{eqn:local_update_grad}
\eea
Here, $\Neighbors$ is the neighbor set for a node in the mini-batch $\Minibatch$. $\Neighbors$ is another mini-batch of nodes sampled from \Vertices. Note that $|\Neighbors| \ll |\Vertices|=N$.

### 2017-07-18 03-sequential-evaluation.tex
Process comments by Kees V and Ceriel (but don't oblige some of them)

In all implementations, \textit{update\_phi} dominates the computation.
The translation from Python into 'Python Idiom' C++ speeds up this stage
by a factor~171. % , even though the Python implementation uses Numpy for its data
% structures. 
Our C++ optimizations speed up this stage
by another factor of~6. Reducing the floating-point precision to 32~bits
doubles this factor, which again shows that the algorithm is very much data-dominant.
%
\textit{update\_pi} has a disproportionately large speed
gain after removing inefficiencies. The Numpy implementation updated the full
$\pi$ array (incidentally, rather efficiently), whereas our C++ optimization
limits the updates to only those values of $\pi$ that have changed.
%
\textit{update\_theta\_beta} gave fewer opportunities for C++ optimization,
in the sampling stages there was none. In \textit{mini-batch-sampling} there
is no speedup compared to Python because it consists of a call to
a fast Python random primitive that cannot be bettered from C++.
The total performance gain between Python and the 64-bit float baseline C++
implementation exceeds a factor of~1000, and this grows to a factor of over~1500
if the floating-point precision is reduced to 32~bits. The 32-bit precision
implementation will be the baseline for the parallel performance comparison.

@@

A multi-threaded version was created by annotating loops in the
baseline implementation with OpenMP primitives in a straightforward
fashion. This accelerated the stages \textit{update\_phi} and
\textit{update\_pi} with a factor~4 for the very small problem instance
in Table~\ref{table-python-comparison}. For larger problem instances, the
speedup for these dominant phases grew to a factor~10.
% The vectorizing OpenCL
% backend described in the next Section again significantly improves the
% performance, but at the cost of substantial coding effort.
% \textbf{\emph{Question: speak about the other stages? they are pipelined.}}

>>>>

In all implementations, \textit{update\_phi} as given in Equations \ref{eqn:local_update}
and \ref{eqn:local_update_grad} dominates the computation.
The translation from Python into 'Python Idiom' C++ speeds up this stage
by a factor~171. % , even though the Python implementation uses Numpy for its data
% structures. 
Our C++ optimizations speed up this stage
by another factor of~6. Reducing the floating-point precision to 32~bits
doubles this factor, which again shows that the algorithm is very much data-dominant.
%
\textit{update\_pi} has a disproportionately large speed
gain after removing inefficiencies. The Numpy implementation updated the full
$\pi$ array (incidentally, rather efficiently), whereas our C++ optimization
limits the updates to only those values of $\pi$ that have changed.
%
\textit{update\_theta\_beta} gave fewer opportunities for C++ optimization,
in the sampling stages there was none. In \textit{mini-batch-sampling} there
is no speedup compared to Python because it consists of a call to
a fast Python random primitive that cannot be bettered from C++.
The total performance gain between Python and the 64-bit float baseline C++
implementation exceeds a factor of~1000, and this grows to a factor of over~1500
if the floating-point precision is reduced to 32~bits. The 32-bit precision
implementation will be the baseline for the parallel performance comparison.

@@

A multi-threaded version was created by annotating loops in the
baseline implementation with OpenMP primitives in a straightforward
fashion. This accelerated the stages \textit{update\_phi} and
\textit{update\_pi} with a factor~4 for the very small problem instance
in Table~\ref{table-python-comparison}. For larger problem instances, the
speedup for these dominant phases grew to a factor~10 on a DAS5 machine.
% The vectorizing OpenCL
% backend described in the next Section again significantly improves the
% performance, but at the cost of substantial coding effort.
% \textbf{\emph{Question: speak about the other stages? they are pipelined.}}

### 2017-07-18 06-conclusion.tex
Process comments by Kees V and Ceriel (but don't oblige some of them)

After a thorough
analysis of the computational patterns and data structures,
we showed that the algorithm's state can be reduced by 75\%,
thus significantly
reducing its data intensity.

@@

We further created a highly scalable implementation to handle the largest existing
community graphs on a distributed cluster machine. The most important data structures
are partitioned across the collective memory of the cluster nodes, which
significantly hurts data access latency and bandwidth for our already data-dominant algorithm.
We made several contributions to improve performance and still maintain scalability:
we created a custom distributed Key-Value store implemented with high-speed RDMA
primitives, that exploits the algorithm's structure to avoid concurrency issues; we
use MPI for synchronization and work distribution between the master and worker nodes;
and
we overlapped computation
with communication to hide latency.
%
We argued that it is sensible to use multicore CPUs in stead of GPUs in this
distributed setting, because GPUs require staging the data piecemeal into GPU
memory, thus counter-balancing any speed gain. Moreover, the computation time is
fully hidden by the latency of remote data accesses.
%
Our empirical evaluation using 65 cluster nodes and large data sets
shows that the system exhibits good strong and weak scalability. We also compared its
performance against a 40-core shared-memory system with 1TB of memory.
The distributed system has superior performance because it has more cores and
more collective memory bandwidth, even though it must fetch its data through the network.
% Additionally, we assessed the efficiency of the algorithm's resource utilization.
Finally, we provide a
demonstration of the implementation's utility by processing 6
large real-world data sets.
To the best of
our knowledge, this is the first time that the problem of deducing overlapping
communities has been learned for problems of such a large scale.

>>>>

After a thorough
analysis of the computational patterns and data structures,
we showed that the algorithm's state size can be reduced by 75\%,
thus significantly
reducing its data intensity.

@@

We further created a highly scalable implementation to handle the largest existing
community graphs on a distributed cluster machine. The most important data structures
are partitioned across the collective memory of the cluster nodes, which
significantly hurts data access latency and bandwidth for our already data-dominant algorithm.
We made several contributions to improve performance and still maintain scalability:
we created a custom distributed Key-Value store implemented with high-speed RDMA
primitives, that exploits the algorithm's structure to avoid concurrency issues; we
use MPI for synchronization and work distribution between the master and worker nodes;
and
we overlapped computation
with communication to hide latency.
%
We argued that it is sensible to use multicore CPUs in stead of GPUs in this
distributed setting, because GPUs require piecemeal staging of the data into GPU
memory, thus counter-balancing any speed gain. Moreover, the computation time is
fully hidden by the latency of remote data accesses.
%
Our empirical evaluation using 65 cluster nodes and large data sets
shows that the system exhibits good strong and weak scalability. We also compared its
performance against a 40-core shared-memory system with 1TB of memory.
The distributed system has superior performance because it has more cores and
more collective memory bandwidth, even though it must fetch its data through the network.
% Additionally, we assessed the efficiency of the algorithm's resource utilization.
Finally, we provide a
demonstration of the implementation's utility by processing 6
large real-world data sets.
To the best of
our knowledge, this is the first time that the problem of deducing overlapping
communities has been learned for problems of such a large scale.

### 2017-07-18 gpu-04-03-accelerator.tex
Process Alex Uta's comments
Comments Henri on what's new&unpublished
Process comments by Kees V and Ceriel (but don't oblige some of them)

\subsection{Design of GPU Optimizations}

@@

The algorithm relies on a set data structure to store the edges of the graph.
This set is queried frequently with randomly generated edges to check for their
membership. In order to improve the performance of such lookups, a custom set
implementation was developed that restricts the set's features based on an
observation of its usage patterns.
The graph is immutable, hence its maximum
size can be specified at creation time and need not be adjustable or provide
thread-safety guarantees. The set data
structure is built as a variant of a cuckoo
hash~\cite{Pagh:2004:CH:1006424.1006426} where each element is a 64-bit
field containing a tuple of the two 32-bit vertices. The cuckoo hash
uses two hash functions to index their corresponding spaces. Additionally, the
cuckoo hash supports upt to 4~keys per bucket. This hash design allows for a load
factor upwards of 90\%, a very economical memory usage.
%% FIXME: Talk lookup hit/miss with 4 keys per bucket?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Parallelization \& Data Dependencies}
% Division into kernels
The original algorithm was structurally reorganized into 4 main sections, each with
one or more kernels, data dependencies and synchronization requirements.
%
First, the sampling of a mini-batch of edges is done on the host as it is a
cheap operation. The mini-batch sampling is followed by the neighbor sampling
kernel which generates uniformly random neighbors for each vertex in the
mini-batch. 
%
Second, the predominant kernel, \textit{update\_phi} is invoked to calculate the
gradients for each vertex $i$ in the mini-batch \Minibatch and update the value of $\phi_i
\mid i \in \Minibatch$. Next, the \textit{update\_pi} kernel is invoked to normalize the
individual $\phi_i$ and store the result in the corresponding~$\pi_i$.
%
Third, the kernels \textit{update\_theta} and \textit{update\_beta} are invoked to modify the
global parameters.
%
Finally, perplexity calculation is performed in a dedicated kernel.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Single floating point vs double
Another modification to the original algorithm is migrating all of its
computation from double to single precision floating-point, analogous to the change
for the baseline implementation in Section~\ref{sec-seq-performance}.
%
This change affects the algorithm in multiple ways.
It again halves the memory
footprint of the algorithm and frees registers.
Besides, on our GPUs it
reduces
the computation intensity besides the data intensity.

>>>>

\subsection{Design of GPU Optimizations}
\label{sec-gpu-design}

@@

The algorithm relies on a set data structure to store the edges of the graph.
This set is queried frequently with randomly generated edges to check for their
membership. In order to improve the performance of such lookups, a custom set
implementation was developed that restricts the set's features based on an
observation of its usage patterns.
The graph is immutable, hence its maximum
size can be specified at creation time and need not be adjustable or provide
thread-safety guarantees. The set data
structure is built as a variant of a cuckoo
hash~\cite{Pagh:2004:CH:1006424.1006426} where each element is a 64-bit
field containing a tuple of the two 32-bit vertices. The cuckoo hash
uses two hash functions to index their corresponding spaces. Additionally, the
cuckoo hash supports up to 4~keys per bucket. This hash design allows for a load
factor upwards of 90\%, a very economical memory usage.
%% FIXME: Talk lookup hit/miss with 4 keys per bucket?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Parallelization and Data Dependencies}
% Division into kernels
The original algorithm was structurally reorganized into 4 main sections, each with
one or more kernels, data dependencies and synchronization requirements.
%
First, the sampling of a mini-batch of edges is done on the host as it is a
cheap operation. The mini-batch sampling is followed by the neighbor sampling
kernel which generates uniformly random neighbors for each vertex in the
mini-batch. 
%
Second, the predominant kernel, \textit{update\_phi} is invoked to calculate the
gradients for each vertex $i$ in the mini-batch \Minibatch and update the value of $\phi_i
\mid i \in \Minibatch$. Next, the \textit{update\_pi} kernel is invoked to normalize the
individual $\phi_i$ and store the result in the corresponding~$\pi_i$.
%
Third, the kernels \textit{update\_theta} and \textit{update\_beta} are invoked to modify the
global parameters.
%
Finally, perplexity calculation is performed in a dedicated kernel.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Single floating point vs double
Another modification to the original algorithm is migrating all of its
computation from double to single precision floating-point, analogous to the change
for the baseline implementation in Section~\ref{sec-seq-performance}.
%
This change affects the algorithm in multiple ways.
It again halves the memory
footprint of the algorithm and frees registers.
Besides, on our GPUs it
reduces
the computation intensity as well as the data intensity.

### 2017-07-18 gpu-05-05-perplexity.tex
Process comments by Kees V and Ceriel (but don't oblige some of them)

This section presents the evolution of the perplexity, i.e.\ the metric for
the solution quality, as a function of time for
two datasets, whose properties are given in Table~\ref{table-snap}. This
analysis shows that our implementations of the algorithm indeed achieve
convergence, and it gives an impression of the time required to achieve
convergence for the largest datasets and community sizes that fit into the memory
of a GPU. The dominant customer of memory is the
$\pi$ matrix, sized $N{\times}K$ for $K$ communities and $N$ vertices in the
graph. The runtime per iteration is independent of~$N$, but it depends on the
product of~$K$, the mini-batch size~$M$, and the neighbor sample size~$|\Neighbors|$. On
the other hand, the number of iterations required to achieve sufficient
convergence does grow with the size of the graph.

Figure~\ref{fig-ppx-gpu-cpu}(a) shows the perplexity development for the dataset
com-DBLP, with $N=317,080$. It is computed on the Maxwell \mbox{Titan-X} GPU with input
parameter set $K=4096$, $M=4096$, $|\Neighbors|=32$. Convergence is reached after 6~hours.
%
Note that the GTX~\mbox{Titan-X} would be unable to
process more than $K$=256 communities for the much larger com-LiveJournal dataset.

>>>>

This section presents the evolution of the perplexity, i.e.\ the metric for
the solution quality, as a function of time for
the DBLP dataset whose properties is given in Table~\ref{table-snap}. This
analysis shows that our implementation of the algorithm indeed achieves
convergence, and it gives an impression of the time required to achieve
convergence for the largest datasets and community sizes that fit into the memory
of a GPU. The dominant customer of memory is the
$\pi$ matrix, sized $N{\times}K$ for $K$ communities and $N$ vertices in the
graph. The runtime per iteration is independent of~$N$, but it depends on the
product of~$K$, the mini-batch size~$M$, and the neighbor sample size~$|\Neighbors|$. On
the other hand, the number of iterations required to achieve sufficient
convergence does grow with the size of the graph.

Figure~\ref{fig-ppx-gpu-cpu} shows the perplexity development for the dataset
com-DBLP, with $N=317,080$. It is computed on the Maxwell \mbox{Titan-X} GPU with input
parameter set $K=4096$, $M=4096$, $|\Neighbors|=32$. Convergence is reached after 6~hours.
%
Note that the GTX~\mbox{Titan-X} would be unable to
process more than $K$=256 communities for the much larger com-LiveJournal dataset.
